The Galactic Midplane is Not a Plane: Implications for {\em Gaia}

### 2019-04-22 paper/ms.tex
wording changes to abstract and intro from meeting discussion

Stellar actions, computed for stars from their six-dimensional phase space measurement and
an assumed Galactic potential, are used to label and distinguish orbits. In
principle, actions have the attractive quality of being invariant and thus
good labels of a star's orbit. 
%mm Of course,
    Such analyses are often used to interpret data from the {\em Gaia}
    mission. However, 
inaccurate measurements of the 
phase space position and uncertainties in the assumed Galactic potential
induce an error in the computed action. We show that, in addition to these
complications, a systematic bias in the phase space position induces a
phase-dependence in the actions, which we interpret as 
%mm an 
    a systematic
error. The induced error distribution is
non-Gaussian and bimodal, 
with neither of the modes peaking on
%mm  the true action value. [the error in a Gaussian peaks on zero] 
    the null value.
A midplane error of $\sim30\,\pc$ for a thin disk orbit or
$\sim225\,\pc$ for a thick disk orbit is 
%mm all that is necessary 
      sufficient
to induce a $100\,\%$ error in the vertical action $J_z$.
%mm [fragment?] orbits, respectively. 
Furthermore, we show that the local midplane varies by
$\sim100\,\pc$ at the Solar circle in Milky Way-mass cosmological zoom-in
simulations from the FIRE collaboration. Thus, current state-of-the-art action
calculations --- which assume the global and local midplanes are the same ---
include a systematic 
%mm offset in $z$.
   vertical offset variation. 
Variation in the local standard of rest
%mm will induce 
     induces
similar issues. The variation of the midplane must be taken into
account when performing dynamical modelling across
the large regions of the disk
accessible 
%mm with the {\em Gaia} mission. [moved first mention up to top of
%abstract, given that this is in the title.]
    to {\em Gaia}.

@@

Once a Galactocentric coordinate system has been established and a
six-dimensional (6D) phase
space measurement of a star has been made, it is often desirable to convert
this measurement into action space to concisely summarize its projected orbit,
model the stellar distribution function, or find stars with similar dynamical
properties. Actions are conserved quantities that describe the orbit of a
star under the assumption of regular, bound orbits in a system where the
equations of motion are separable in a particular coordinate system. They are
the 
%mm typical
integral of the Hamiltonian in cylindrical coordinates 
%mm and are given by,
\beq\label{eq:actions}
J_i \equiv
\frac{1}{2\pi} \oint_{\text{orbit}}p_i\,\text{d}x_i\text{,}
\eeq
where $i=R,z,\phi$ are the radial, vertical, and azimuthal coordinates
respectively in a cylindrical coordinate system and the $p_i$ are the
conjugate momenta.\footnote{We 
%mm typically [there are no other J_R's in manuscript]
write $J_r$ instead of $J_R$ for
aesthetic reasons.} In a slowly-evolving axisymmetric potential, these actions
are invariant and $J_{\phi} \equiv L_z$
\citep{2008gady.book.....B,2014RvMP...86....1S}.

>>>>

Stellar actions, computed for stars from their six-dimensional phase space measurement and
an assumed Galactic potential, are used to label and distinguish orbits. In
principle, actions have the attractive quality of being invariant and thus
good labels of a star's orbit. 
%mm Of course,
    Such analyses are often used to interpret data from the {\em Gaia}
    mission. However, 
inaccurate measurements of the 
phase space position and uncertainties in the assumed Galactic potential
induce an error in the computed action. We show that, in addition to these
complications, a systematic bias in the phase space position induces a
phase-dependence in the actions, which we interpret as 
%mm an 
    a systematic
error. The induced error distribution is
non-Gaussian and bimodal, 
with neither of the modes peaking on
%mm  the true action value. [the error in a Gaussian peaks on zero] 
    the null value.
A variation in the vertical position midplane of $\sim30\,\pc$ for a thin disk orbit or
$\sim225\,\pc$ for a thick disk orbit is 
%mm all that is necessary 
      sufficient
to induce a $100\,\%$ systematic error in the vertical action $J_z$ computed under the assumption of an axisymmetric potential.
%mm [fragment?] orbits, respectively. 
Furthermore, we show that the local midplane varies by
$\sim100\,\pc$ at the Solar circle in Milky Way-mass cosmological zoom-in
simulations from the FIRE collaboration. Thus, current state-of-the-art action
calculations --- which assume the global and local midplanes are the same ---
include a systematic 
%mm offset in $z$.
   vertical offset variation. 
Variation in the local standard of rest
%mm will induce 
     induces
similar issues. The variation of the midplane must be taken into
account when performing dynamical modelling across
the large regions of the disk
accessible 
%mm with the {\em Gaia} mission. [moved first mention up to top of
%abstract, given that this is in the title.]
    to {\em Gaia}.

@@

Once a Galactocentric coordinate system has been established and a
six-dimensional (6D) phase
space measurement of a star has been made, it is often desirable to convert
this measurement into action space to concisely summarize its projected orbit,
model the stellar distribution function, or find stars with similar dynamical
properties. Actions are conserved quantities that describe the orbit of a
star under the assumption of regular, bound orbits in a system where the
equations of motion are separable in a particular coordinate system. They are
the 
%mm typical
cyclical integral of the canonical momentum  over its conjugate position,
%mm and are given by,
\beq\label{eq:actions}
J_i \equiv
\frac{1}{2\pi} \oint_{\text{orbit}}p_i\,\text{d}x_i\text{,.}
\eeq
Under the assumption of axisymmetry, $i=R,z,\phi$ are the radial, vertical, and azimuthal coordinates
respectively in a cylindrical coordinate system and the $p_i$ are the
conjugate momenta.\footnote{We 
%mm typically [there are no other J_R's in manuscript]
write $J_r$ instead of $J_R$ for
aesthetic reasons.} In a slowly-evolving axisymmetric potential, these actions
are invariant and $J_{\phi} \equiv L_z$
\citep{2008gady.book.....B,2014RvMP...86....1S}.
